Improving the reliability of cognitive task measures: \break A narrative review

### 2023-01-09 manuscript/main.tex
manuscript revisions + letter to reviewers

% load packages
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[a4paper, margin=0.8in]{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage[backend=biber,style=nature]{biblatex}

@@

@@

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1,*]{Samuel Zorowitz}
\author[1,2]{Yael Niv}
\affil[1]{Princeton Neuroscience Institute, Princeton University, USA}
\affil[2]{Department of Psychology, Princeton University, USA}
\affil[*]{Corresponding author {\small (zorowitz@princeton.edu)}}

@@

\abstract{Cognitive tasks are capable of providing researchers with crucial insights into the relationship between cognitive processing and psychiatric phenomena across individuals. However, many recent studies have found that task measures exhibit poor reliability, which hampers their utility for individual-differences research. Here we provide a narrative review of approaches to improve the reliability of cognitive task measures. First, we review methods of calculating reliability and discuss some nuances that are specific to cognitive tasks. Then, we introduce a taxonomy of approaches for improving task reliability. Where appropriate, we highlight studies that are exemplary for improving the reliability of specific task measures. We hope that this article can serve as a helpful guide for experimenters who wish to design a new task, or improve an existing one, to achieve sufficient reliability for use in individual-differences research.}

@@

% P1: Using valid and reliable measures of behavior is crucial for advancing biological psychiatry.

Cognitive tasks have already demonstrated, and continue to hold, tremendous promise for biological psychiatry. Individual differences in performance on cognitive tasks provide researchers with crucial insights into the relationship between cognitive processing and behavioral phenomena in the clinic and in everyday life (e.g. reward sensitivity in a monetary-incentivized task and anhedonia in depression). Elsewhere in psychology, cognitive tasks have been useful in predicting important outcomes such as academic achievement \cite{spiegel2021relations} and cognitive decline \cite{hartshorne2015does}. Cognitive tasks, then, are invaluable tools for building and refining our understanding of psychiatric symptoms and syndromes. For a cognitive task to be useful in this regard, however, it must possess sufficient measurement properties.

% P2: Definitions + Three key properties of behavioral measures: discriminability, reliability, and validity

We define a cognitive task as any experimental paradigm that measures behavior in order to make inferences about one or more cognitive processes (e.g. Stroop task, delay discounting task, reversal learning task). Cognitive-task measures for characterizing individual differences are descriptive or model-based indicators of behavioral performance such as proportion correct, average response time, and learning rate. The psychometric quality of such a measure can be summarized by three key properties: \emph{discriminating power}, \emph{validity} and \emph{reliability} \cite{kline2015handbook}. The discriminatory power of a task measure describes its ability to measure variability in participants' performance. This is a necessary property of tasks to be used for studying individual differences; where there is no variation in performance, there are no individual differences to study. The validity of a task measure concerns whether it actually measures what it intends to measure. Finally, the reliability of a task measure characterizes the degree to which it consistently rank-orders participants across measurement instances. Put another way, a task measure is reliable if, assuming participants have not changed, it produces the same score for each participant over time. The focus of this review is task reliability. 
% P3: definition and Importance of reliability
\subsubsection*{The formal definition of reliability}

In classical test theory, the variance in observed scores on a task measure $x$ is the sum of the true score variance $\sigma^2_T$, reflecting real individual differences in the latent construct of interest, and measurement error $\sigma^2_E$, i.e., $\sigma^2_x = \sigma^2_T + \sigma^2_E$. The reliability of a measure is defined as the proportion of variance attributable to the true score variance relative to total variance: $\rho_{xx'} = \frac{\sigma^2_T}{\sigma^2_T + \sigma^2_E}$. Thus, reliability quantifies the magnitude of individual differences relative to the noisiness of a task measure; the larger the reliability of a task measure, the more it reflects true individual differences rather than reflecting noise. Reliability is therefore a prerequisite for validity: an unreliable task measure reflects measurement error and not the construct of interest. If this were not reason enough to care about reliability, the observed correlation between two measures (e.g. task performance and self-reported symptom score) is bounded by their individual reliabilities \cite{Spearman1904-mo}: 
\begin{equation}
    \rho_{xy} = \rho_{tt} \sqrt{\rho_{xx'} \cdot \rho_{yy'}}
\end{equation}
where $\rho_{xx'}$ and $\rho_{yy'}$ are the reliabilities of two measures, $x$ and $y$; $\rho_{tt}$ is their true latent correlation; and  $\rho_{xy}$ is their observed correlation. As all reliabilities are $<1$, the reliability of a measure places an upper bound on the maximum observable correlation between itself and a second measure (Figure \ref{fig:fig01}). As an important corollary, as measure reliability decreases, the number of participants required to reliably detect a correlation between two measures increases \cite{Parsons2019-jw}. Thus, poor reliability hampers our ability to investigate associations between cognitive processes, as measured by task performance, and other variables of interest.

@@

% P4: Reliability is inherently context-dependent: it is a function of the population, task, and testing context

To further complicate matters, the reliability of a task measure is not absolute -- it reflects interactions between task design, the participants, and the context in which the task is administered. Indeed, task reliability can vary as a function of parameters of the experiment such as the specific stimulus set, number of trials, time limits \cite{paap2016role, cooper2017role}; sample populations (healthy adults, children, psychiatric patients \cite{arnon2020current, cooper2017role}); testing locations (in clinic, online); response modality (desktop, smartphone, virtual reality \cite{pronk2022can, bruder2021reliability}); and scoring or estimation method \cite{Rouder2019-am, haines2020learning}. For example, a cognitive task originally designed for use with an adult population may prove too difficult for children. Their task performance may drop to chance level, thereby minimizing between-participant variance and, as a consequence, task reliability. As a second example, participants completing an experiment online from their homes may experience more distraction than if they participated in the lab. This may increase measurement error, leading to a concomitant decrease in reliability. Experimenters therefore cannot assume the reliability of a task measure is constant. Ideally researchers would investigate and report the reliability of task measures as part of any individual differences research or, at the very least, after having made changes to a task or scoring procedure or when administering the task to new sample populations or in new testing contexts.

% P5: Task reliability lower on average

@@

One possible explanation for this finding is the so-called ``reliability paradox'' of cognitive tasks \cite{Hedge2018-lf}, which states that the often lackluster reliability of tasks is a result of a mismatch in goals between experimental and individual-differences psychological research. In experimental psychology, the goal is often to demonstrate the existence of a behavioral effect. One means of increasing the power to detect an effect is to minimize between-participants variance. This is the exact opposite of what is desirable for individual differences research, where between-participants variance is essential to achieving reliable task measures. Thus, the tendency in biological psychiatry to adopt the most prominent tasks in experimental psychology---the ones that most reliably demonstrate a behavioral effect---may actually hamstring our efforts to study individual differences.

% P6: Purpose of paper: narrative review of task reliability

Regardless, we do not believe that task measures are inherently less reliable than self-report measures, and that pessimism about task-based individual-differences research is warranted. It is possible to (re)design tasks to achieve good reliability, even to the high levels dictated by conventional standards \cite{waltmann2022sufficient, sullivan2022enhancing, kucina2022solution, snijder2022psychometric}. The purpose of the current article is to provide a narrative review of approaches to improve task-measure reliability. The article is divided into two main parts. First, we review methods of calculating reliability and discuss some nuances that are specific to cognitive tasks. Then, we introduce a taxonomy of approaches for improving task reliability using concrete examples from the published literature. We hope that this article can serve as a helpful guide for experimenters who wish to design a new task, or improve an existing task, to achieve sufficient reliability for use in individual-differences research.

@@

There are two measures of reliability of relevance for task measures: internal consistency and test-retest reliability. Internal consistency is the reliability of a measure in a single administration of a task. Test-retest reliability (also known as temporal stability) is the reliability of a task measure across two or more administrations. Even for cross-sectional experiments, where stability over time is of no scientific interest, test-retest reliability is a useful index. This is because estimates of internal consistency tend to be inflated due to between-participants variance from construct-irrelevant, state-dependent factors (e.g. current mood, fatigue). Given sufficient time between testing sessions, test-retest reliability should be less biased by state factors, and thus a better estimate of the true cross-sectional task-measure reliability. However, whereas high internal consistency is always desirable for individual difference studies, high test-retest reliability may not be, depending on the construct a researcher is intending to measure. For example, low one-month test-retest reliability may not be problematic for an index of a transient cognitive process (e.g., mood-dependent attentional biases), but is a problem if individual differences in a cognitive process are hypothesized to be stable across time (e.g., extraversion and social reward processing).

@@

Calculating the internal consistency of a task measure is more complicated. The most common measure of internal consistency, Cronbach's $\alpha$, requires certain assumptions that are unrealistic for many task data (e.g., equivalence of trials, uncorrelated measurement error). As such, internal consistency for task measures is instead usually calculated via split-$K$ reliability, where $K$ indicates the number of splits. Split-half reliability ($K=2$) is most common. A critical challenge in calculating split-half reliability is in deciding how to partition the data, as estimates of reliability may be biased when partitions are not equally good measures of a behavior \cite{Green2016-xw, pronk2022can}. For example, first-second splitting (i.e., partitioning the data into the first and second halves of an experiment) may underestimate reliability due to differences between the partitions resulting from practice, fatigue, or other linear time effects. In contrast, odd-even splitting (i.e., partitioning the data into odd and even trials) may be confounded when behavior across trials is non-independent (i.e., measurement error is correlated across trials). Under such conditions, odd-even splitting will artificially inflate the similarity of data across partitions, thereby decreasing estimates of measurement noise and overestimating reliability. Therefore, where possible, a permutation-based approach to calculating split-half reliability is recommended \cite{pronk2022can, Parsons2019-jw}. Here, reliability is averaged across many thousands of random partitions of the data into halves. For task measures derived from cognitive models, however, it may be prohibitively computationally intensive to employ such an approach due to the need to re-estimate the model for each new subset of the data. Moreover, for learning tasks that are commonly used in computational psychiatry, cognitive-model based task measures cannot be estimated from only a subset of the trials due to inherent non-independence of task behavior across trials. In these cases, one can design tasks with at least two independent blocks. The model can then be fit to each block independently and reliability calculated using the model parameters estimated from each.

As a final point, traditional sum or mean score estimates of performance (e.g., proportion correct responses, mean response time) may substantially underestimate task reliability \cite{Rouder2019-am, haines2020learning}. This is because such summary scores are contaminated by trial-level noise that, in the absence of sufficiently large number of trials, increases the measurement error (and thus diminishes reliability). Instead, it is often preferable to use generative or statistical models that account for trial-level variability in behavior. By directly modeling trial-by-trial variability, estimates of participants' performance are effectively de-noised. It is also advisable to use hierarchical models that partially pool data across participants, which in turn further increases the precision of estimates of participants' performance. Together, these methods can substantially boost estimates of reliability \cite{Rouder2019-am, haines2020learning}, as multiple studies have demonstrated  \cite{snijder2022psychometric, sullivan2022enhancing, brown2020improving, waltmann2022sufficient}. For a detailed discussion of generative models in the context of task reliability, see Haines (this issue).

@@

By definition, the reliability of a task measure is zero when there is no variability across participants. Thus, range restriction of task measures via ceiling or floor effects is a serious obstacle to reliability. Siegelman and colleagues \cite{siegelman2017measuring} noted the consequences of floor effects on reliability in the context of statistical learning tasks. In such tasks, participants must learn to identify subtle patterns in the transition probabilities underlying a continuous sequence of stimuli. In reanalyzing archival datasets, Siegelman and colleagues found a majority of participants were at chance-level performance in discriminating between legitimate and foil sequence patterns; consequently, the reliability of conventional proportion correct measures suffered. In response, the authors designed a new statistical learning task involving stimulus sequences that ranged more widely in their difficulty to learn. Only a minority of participants showed chance-level performance on this new task and, as such, the reliability of proportion correct scores improved. Similarly, in developing an abbreviated working-memory task, Oswald and colleagues \cite{oswald2015development} found that they could remove the easiest trials---those with ceiling level performance---with virtually no change to task reliability. This is because those trials are incapable of differentiating ability across participants and therefore cannot contribute significantly to the reliability of the task.

Researchers administering a task to a new population should be especially wary of range-restriction effects. Cognitive tasks calibrated for one group of participants may not be adequately sensitive for others due to being too easy or difficult for a different group. For example, Arnon and colleagues \cite{arnon2020current} found that statistical learning tasks developed for adults were too difficult for young children and therefore yielded unreliable discrimination scores in that population. Similarly, Kyllonen and colleagues \cite{kyllonen2019general} were motivated to develop a battery of fluid-reasoning measures for highly educated adults due to observed ceiling effects in performance when using preexisting fluid-reasoning tasks in this population.

@@

A related issue for task reliability is practice effects, where participants' performance on a task improves with repeated administrations. Practice effects are relatively common for cognitive tasks \cite{hausknecht2007retesting, scharfen2018retest}. They might occur due to the attenuation of task-irrelevant nuisance factors (e.g. performance anxiety) and/or the learning of task-specific knowledge or strategies. Practice effects are not inherently an issue for reliability --- especially if an experimenter is only interested in the consistency, but not the absolute agreement, of participants' performance over time --- but they can become a pernicious issue if they are severe enough to induce ceiling effects. For example, Paredes and colleagues \cite{paredes2021psychometric} observed large practice effects on the Pavlovian go/no-go task for short retest intervals (3 days, 14 days), which resulted in poor estimates of test-retest reliability. In developmental and lifespan studies, practice effects are potentially complicated by their interaction with age \cite{anokhin2022age, salthouse2010influence}; that is, practice-induced ceiling effects may present in some age groups but not others.

@@

A second strategy is to use a combination of clear instructions and practice trials to help participants reach stable performance from the start of an experiment (see the ``Improving experiment designs'' section below). Another strategy is to design tasks so as to prevent or discourage the formation of task-specific strategies. For example, McLean and colleagues \cite{mclean2018towards} investigated the repeatability of the beads task. In the beads task, participants are presented with two jars containing beads of two colors in equal but opposite ratios. In each trial of the task, a predetermined sequence of beads is drawn from one jar. Participants must decide which jar beads were being drawn from or request to see more beads. In a typical version of the task, the same sequence of bead draws is used across all trials. McLean and colleagues found that participants were aware that the sequence repeated across trials and, as a consequence, became more erratic in their decision to witness more bead draws with additional trials. In response, the authors developed a new version of the task that included distractor sequences of bead draws. This new design was effective in preventing participants from becoming aware of the target sequence, which in turn resulted in more consistent responding, which improved the reliability of participants' information seeking scores.

\subsubsection{Enhancing experimental effects}

The preceding sections described potential threats to between-participants variability, but not approaches to improve it. A primary strategy for increasing between-participants variability is to enhance the experimental effect. Amplifying the magnitude of an experimental effect (e.g., making a task more challenging, increasing the potency of affect induction) typically increases the range of participants' responses to it. For example, Kucina and colleagues \cite{kucina2022solution} investigated the reliability of cognitive conflict effects in new versions of several standard cognitive-control tasks (e.g., Stroop, Flanker, Simon) that amplified cognitive interference via two task design features. First, they combined multiple sources of cognitive interference in the same task, for example by combining the Stroop and Simon effects to create a ``Stroopon'' task. Second, for a subset of trials, they required participants to make multiple responses based on both relevant and irrelevant stimuli attributes. Compared to the standard version of these tasks, these experimental manipulations had the effect of increasing task demands, which resulted in greater between-participants variance, and higher reliability as a consequence. (See also Snijder and colleagues \cite{snijder2022psychometric} for a similar redesign of classic cognitive-control tasks that improved reliability by increasing proactive control demands.)

A related strategy is to calibrate the difficulty of the task to the average ability of the population of interest. For example, consider a task trial with two response options. Given the Bernoulli distribution, the variance in responses on this trial will be maximal when the probability of choosing either response is equal. Thus, aggregating across many trials, between-participants variance is maximized --- and task reliability is improved --- when the difficulty of all items is matched to the average ability of the sample \cite{gulliksen1945relation} (or slightly higher if participants can guess the correct response \cite{lord1952relation, feldt1993relationship}). Of course, this design principle is only helpful to the degree that a researcher knows the average ability level of their participants. If this is unknown or poorly characterized, then it is instead desirable to design a task with trials spanning a uniform range of difficulty.

@@

A final strategy for increasing between-participant variance is to simply recruit more diverse samples. While convenient, undergraduate students from a single university are likely to be relatively homogeneous in their cognitive profiles. It may be worthwhile instead to recruit participants from the community or from an online labor market (e.g., Amazon Mechanical Turk, Prolific Academic, CloudResearch Panels). With regard to the latter, because online participants typically complete experiments from their homes or other poorly-controlled environments, they are more likely to be distracted or to multi-task during an experiment \cite{newman2021data}. Thus, when recruiting online samples, experimenters should take special care to ensure that an increase in between-participants variance is not offset by a concomitant increase in measurement noise.

@@

We thank Rachel Bedder for helpful feedback on this manuscript. This project was made possible with support from the National Center for Advancing Translational Sciences (NCATS), a component of the National Institute of Health (NIH), under award number UL1TR003017a and a National Science Foundation Graduate Research Fellowship.

>>>>

% load packages
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[a4paper, margin=0.8in]{geometry}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage{hyperref}           % format hyperlinks
\usepackage[pagewise]{lineno}   % add line numbers
\usepackage{xcolor,soul}        % highlight formatting

@@

% define highlight style
\colorlet{usercolorname}{yellow!00}    % use for public version
% \colorlet{usercolorname}{yellow!30}    % use for reviewer version
% \linenumbers                           % use for reviewer version
\sethlcolor{usercolorname}

@@

% set up the author block
\setlength{\affilsep}{2em}   % set the space between author and affiliation
\author[1,*]{Samuel Zorowitz}
\author[1,2,*]{Yael Niv}
\affil[1]{Princeton Neuroscience Institute, Princeton University, USA}
\affil[2]{Department of Psychology, Princeton University, USA}
\affil[*]{Corresponding authors (zorowitz@princeton.edu, yael@princeton.edu)}

@@

\abstract{Cognitive tasks are capable of providing researchers with crucial insights into the relationship between cognitive processing and psychiatric phenomena. However, many recent studies have found that task measures exhibit poor reliability, which hampers their usefulness for individual-differences research. Here we provide a narrative review of approaches to improve the reliability of cognitive task measures. First, we review methods of calculating reliability and discuss some nuances that are specific to cognitive tasks. Then, we introduce a taxonomy of experiment design and analysis strategies for improving task reliability. Where appropriate, we highlight studies that are exemplary for improving the reliability of specific task measures. We hope that this article can serve as a helpful guide for experimenters who wish to design a new task, or improve an existing one, to achieve sufficient reliability for use in individual-differences research.}

@@

\hl{Cognitive tasks hold great promise for biological psychiatry.} When properly designed, such tasks are capable of isolating and measuring specific cognitive processes. Individual differences in performance on cognitive tasks can therefore provide researchers with crucial insights into the cognitive processes underlying psychiatric phenomena. Elsewhere in psychology, cognitive tasks have been useful in predicting important outcomes such as academic achievement \cite{spiegel2021relations} and cognitive decline \cite{hartshorne2015does}. \hl{Cognitive tasks, then, have the potential to be invaluable tools for refining our understanding of psychiatric symptoms and syndromes.} For a cognitive task to be useful in this regard, however, it must possess sufficient measurement properties.

We define a cognitive task as any experimental paradigm that measures behavior in order to make inferences about one or more cognitive processes (e.g., Stroop task, delay discounting task, reversal-learning task). Cognitive-task measures of behavioral performance can be descriptive (e.g., proportion correct responses, average response time) or model-based (e.g., drift rate in evidence accumulation models). The psychometric quality of a measure can be summarized by three key properties: \emph{discriminatory power}, \emph{validity} and \emph{reliability} \cite{kline2015handbook}. The discriminatory power of a task measure describes its ability to measure variability in participants' performance. This is a necessary property of tasks used to study individual differences; where there is no variation in performance, there are no individual differences to study. The validity of a task measure concerns whether it actually measures what it intends to measure. Finally, the reliability of a task measure characterizes the degree to which it consistently measures some feature of participants. \hl{That is, a task measure is reliable if, assuming participants have not changed, it produces the same scores, or the same ordering of scores, for participants within a single testing session or across multiple testing sessions.} This review focuses on task-measure reliability.

\subsubsection*{The formal definition of reliability}

In classical test theory \cite{allen2001introduction}, the variance in observed scores on a task measure $x$ is the sum of the true score variance $\sigma^2_T$, reflecting real individual differences in the latent construct of interest, and measurement error $\sigma^2_E$, i.e., $\sigma^2_x = \sigma^2_T + \sigma^2_E$. The reliability of a measure is defined as the proportion of variance attributable to the true score variance relative to total variance: $\rho_{xx'} = \frac{\sigma^2_T}{\sigma^2_T + \sigma^2_E}$. Thus, reliability quantifies the magnitude of individual differences relative to the noisiness of a task measure; the larger the reliability of a task measure, the more it reflects true individual differences rather than noise. Reliability is therefore a prerequisite for validity: an unreliable task measure reflects measurement error and not the construct of interest. If this were not reason enough to care about reliability, the observed correlation between two measures (e.g., task performance and self-reported symptom score) is bounded by their individual reliabilities \cite{Spearman1904-mo}: 
\begin{equation}
    \rho_{xy} = \rho_{tt} \sqrt{\rho_{xx'} \cdot \rho_{yy'}}
\end{equation}
where $\rho_{xx'}$ and $\rho_{yy'}$ are the reliabilities of two measures, $x$ and $y$; $\rho_{tt}$ is their true latent correlation; and  $\rho_{xy}$ is their observed correlation. As all reliabilities are $<1$, the reliability of a measure places an upper bound on the maximum observable correlation between itself and a second measure (Figure \ref{fig:fig01}). As an important corollary, as measure reliability decreases, the number of participants required to reliably detect a correlation between two measures increases \cite{Parsons2019-jw}. Thus, poor reliability hampers our ability to investigate associations between cognitive processes, as measured by task performance, and other variables of interest.

@@

To further complicate matters, the reliability of a task measure is not absolute -- it reflects interactions between the task design, the participants, and the context in which the task is administered. Indeed, task reliability can vary as a function of experiment parameters (specific stimulus set, number of trials, time limits \cite{paap2016role, cooper2017role}); sample populations (healthy adults, children, psychiatric patients \cite{arnon2020current, cooper2017role}); testing locations (in clinic, online); response modality (desktop, smartphone, virtual reality \cite{pronk2022can, bruder2021reliability}); \hl{scoring method (component scores, difference scores);} and estimation method \cite{Rouder2019-am, haines2020learning, chen2021trial}. For example, a cognitive task originally designed for use with an adult population may prove too difficult for children. Their task performance may drop to chance level, thereby minimizing between-participant variance and, as a consequence, task reliability. As a second example, participants completing an experiment online from their homes may experience more distraction than if they participated in the lab. This may increase measurement error, leading to a concomitant decrease in reliability. Experimenters therefore cannot assume the reliability of a task measure is constant. At the very least, researchers should evaluate reliability after having made changes to a task or scoring procedure, or when administering the task to new sample populations or in new testing contexts. Ideally, researchers would investigate and report the reliability of task measures as part of any individual differences research.

@@

One possible explanation for this finding is the so-called ``reliability paradox'' of cognitive tasks \cite{Hedge2018-lf}, which states that the often lackluster reliability of tasks is a result of a mismatch in goals between experimental and individual-differences psychological research. In experimental psychology, the goal is often to demonstrate the existence of a behavioral effect. One means of increasing the power to detect an effect is to minimize between-participants variance. This is the exact opposite of what is desirable for individual differences research, where between-participants variance is essential to achieving reliable task measures. \hl{For example, the Stroop effect is one of the most robust effects in experimental psychology; virtually everyone shows a Stroop effect} \cite{haaf2017developing}. \hl{However, in part due to this fact, between-participants variance on the Stroop effect is often limited} \cite{rouder2019most}. Thus, the tendency in biological psychiatry to adopt the most prominent tasks in experimental psychology---the ones that most reliably demonstrate a behavioral effect---may actually hamstring efforts to study individual differences.

Regardless, we do not believe that task measures are inherently less reliable than self-report measures, or that pessimism about task-based individual-differences research is warranted. It is possible to (re)design tasks to achieve good reliability, even to the high levels dictated by conventional standards \cite{mclean2018towards, kucina2022solution, snijder2022psychometric, rey2019executive}. The purpose of the current article is to provide a narrative review of approaches to improve task-measure reliability. The article is divided into two main parts. First, we review methods of calculating reliability and discuss some nuances that are specific to cognitive tasks. \hl{This section is not intended to be exhaustive, but rather to point to other publications where these topics have already been discussed at length.} Then, we introduce a taxonomy of approaches to improve the reliability of cognitive-task measures through experiment design and analysis, using concrete examples from the published literature. We hope that this article can serve as a helpful guide for experimenters designing a new task, improving an existing task, \hl{or refining their scoring methods} to achieve sufficient reliability for use in individual-differences research.

@@

There are two measures of reliability of relevance for task measures: internal consistency and test-retest reliability. Internal consistency is the reliability of a measure in a single administration of a task. Test-retest reliability (also known as temporal stability) is the reliability of a task measure across two or more administrations. Even for cross-sectional experiments, where stability over time is of no scientific interest, test-retest reliability is a useful index. This is because estimates of internal consistency tend to be inflated due to between-participants variance from construct-irrelevant, state-dependent factors (e.g., current mood, fatigue). Given sufficient time between testing sessions, test-retest reliability should be less biased by state factors, and thus a better estimate of the true cross-sectional task-measure reliability. However, whereas high internal consistency is always desirable for individual difference studies, high test-retest reliability may not be, depending on the construct a researcher is intending to measure. For example, low one-month test-retest reliability may not be problematic for an index of a transient cognitive process (e.g., mood-dependent attentional biases), but is a problem if individual differences in a cognitive process are hypothesized to be stable across time (e.g., extraversion and social reward processing).

@@

Calculating the internal consistency of a task measure is more complicated. \hl{The most common measure of internal consistency is Cronbach's $\alpha$, which is a function of the average correlation across all unique pairs of trials. However, Cronbach's $\alpha$ is an accurate measure of reliability only under assumptions that are unrealistic for many tasks (e.g., equivalence of trials, uncorrelated measurement error;} \cite{Green2016-xw, Parsons2019-jw}). \hl{As such, internal consistency for task measures is instead usually calculated via split-half reliability, where reliability is estimated after trial data have been divided into two halves.} A critical challenge in calculating split-half reliability is in deciding how to partition the data, as estimates of reliability may be also biased if the data partitions violate either of the two above assumptions (for detailed discussion, see \cite{Green2016-xw, Pronk2022-mq}). For example, first-second splitting (i.e., partitioning the data into the first and second halves of an experiment) may underestimate reliability due to nonequivalence of the two partitions resulting from practice, fatigue, or other linear time effects. In contrast, odd-even splitting (i.e., partitioning the data into odd and even trials) may cause bias when behavior across trials is non-independent (i.e., measurement error is correlated across trials),  artificially inflating the similarity of data across partitions and thereby decreasing estimates of measurement noise and overestimating reliability. Therefore, where possible, a permutation-based approach to calculating split-half reliability is recommended \cite{Pronk2022-mq, Parsons2019-jw}. Here, reliability is averaged across many thousands of random partitions of the data into halves. \hl{(Insofar that Cronbach's $\alpha$ is analytically equivalent to the average of all possible split-half reliability estimates} \cite{cronbach1951coefficient}, \hl{permutation-based split-half reliability provides an approximation to Cronbach's $\alpha$ while avoiding its problematic assumptions.)} For task measures derived from cognitive models, however, it may be prohibitively computationally intensive to employ such an approach due to the need to re-estimate the model for each new subset of the data. Moreover, for learning tasks that are commonly used in computational psychiatry, cognitive-model based task measures cannot be estimated from only a subset of the trials due to inherent non-independence of task behavior across trials. In these cases, one can design tasks with at least two independent blocks. The model can then be fit to each block independently and reliability calculated using the model parameters estimated from each.

As a final point, traditional sum or mean score estimates of performance (e.g., proportion correct responses, mean response time) may substantially underestimate task reliability \cite{Rouder2019-am, haines2020learning, chen2021trial}. \hl{This is because such summary scores are contaminated by trial-level noise that, in the absence of a sufficiently (possibly prohibitively) large number of trials, increases measurement error (and thus diminishes reliability). Instead, it may be preferable to use trial-level hierarchical models in which observations are organized hierarchically (e.g., individuals within a group, trials within an individual) with variability modeled at both levels. Hierarchical models exert a pooling or regularization effect on person-level variables, in effect correcting for measurement error and improving estimates of reliability} \cite{Rouder2019-am, haines2020learning, chen2021trial}. The benefits of hierarchical models for estimating reliability has been multiply demonstrated \cite{snijder2022psychometric, sullivan2022enhancing, brown2020improving, waltmann2022sufficient}, \hl{though see} \cite{rouder2019most} \hl{for discussion of when these benefits may be limited. Using statistical models that more accurately characterize the latent data-generating process (e.g., using the shifted log-normal distribution to model response times) may also improve reliability estimates} \cite{haines2020learning, price2019computational}. For a detailed discussion of hierarchical and generative models in the context of task reliability, see Haines and colleagues (this issue; \cite{haines2020learning}).

@@

By definition, the reliability of a task measure is zero when there is no variability across participants. Thus, range restriction of task measures via ceiling or floor effects is a serious obstacle to reliability. Siegelman and colleagues \cite{siegelman2017measuring} noted the consequences of floor effects on reliability in the context of statistical learning tasks. In such tasks, participants must learn to identify subtle patterns in the transition probabilities underlying a continuous sequence of stimuli. In reanalyzing archival datasets, Siegelman and colleagues found a majority of participants were at chance-level performance in discriminating between legitimate and foil sequence patterns; consequently, the reliability of conventional proportion correct measures suffered. In response, the authors designed a new statistical learning task involving stimulus sequences that ranged more widely in their difficulty to learn. Only a minority of participants showed chance-level performance on this new task and, as such, the reliability of proportion correct scores improved \hl{(from $\rho = 0.75$ to $0.88$)}. Similarly, in developing an abbreviated working-memory task, Oswald and colleagues \cite{oswald2015development} found that they could remove the easiest trials---those with ceiling level performance---with virtually no change to task reliability. This is because those trials are incapable of differentiating ability across participants and therefore cannot contribute significantly to the reliability of the task.

Researchers administering a task to a new population should be especially wary of range-restriction effects. Cognitive tasks calibrated for one group of participants may not be adequately sensitive for others due to being too easy or difficult for a different group. For example, Arnon and colleagues \cite{arnon2020current} found that statistical learning tasks developed for adults were too difficult for young children and therefore yielded unreliable discrimination scores in that population. Similarly, Kyllonen and colleagues \cite{kyllonen2019general} developed a battery of fluid-reasoning measures for highly educated adults after observing ceiling effects in performance when using preexisting fluid-reasoning tasks in this population.

@@

A related issue for task reliability is practice effects, where participants' performance on a task improves with repeated administrations. Practice effects are relatively common for cognitive tasks \cite{hausknecht2007retesting, scharfen2018retest}. They might occur due to the attenuation of task-irrelevant nuisance factors (e.g., performance anxiety) and/or the learning of task-specific knowledge or strategies. Practice effects are not inherently an issue for reliability --- especially if an experimenter is only interested in the consistency, but not the absolute agreement, of participants' performance over time --- but they can become a pernicious issue if \hl{they are exhibited differentially across participants} or if they are severe enough to induce ceiling effects. For example, Paredes and colleagues \cite{paredes2021psychometric} observed large practice effects on the Pavlovian go/no-go task for short retest intervals (3 days, 14 days), which resulted in poor estimates of test-retest reliability. In developmental and lifespan studies, practice effects are potentially complicated by their interaction with age \cite{anokhin2022age, salthouse2010influence}; that is, practice-induced ceiling effects may present in some age groups but not others.

@@

A second strategy is to use a combination of clear instructions and practice trials to help participants reach stable performance from the start of an experiment (see the ``Improving experiment designs'' section below). Another strategy is to design tasks so as to prevent or discourage the formation of task-specific strategies. For example, McLean and colleagues \cite{mclean2018towards} investigated the repeatability of the beads task. In the beads task, participants are presented with two jars containing beads of two colors in equal but opposite ratios. In each trial of the task, a predetermined sequence of beads is drawn from one jar. Participants must decide which jar beads were being drawn from or request to see more beads. In a typical version of the task, the same sequence of bead draws is used across all trials. McLean and colleagues found that participants were aware that the sequence repeated across trials and, as a consequence, became more erratic in their decision to witness more bead draws with additional trials. In response, the authors developed a new version of the task that included distractor sequences of bead draws. This new design was effective in preventing participants from becoming aware of the target sequence, which in turn resulted in more consistent responding, which improved the reliability of participants' information seeking scores \hl{(from $\rho = 0.62$ to $0.84$)}.

\subsubsection{Enhancing experimental manipulations}

The preceding sections described potential threats to between-participants variability, but not approaches to improve it. A primary strategy for increasing between-participants variability is to enhance the experimental \hl{manipulation}. Amplifying the strength of an experimental \hl{manipulation} (e.g., making a task more challenging, increasing the potency of affect induction) typically increases the range of participants' responses to it. For example, Kucina and colleagues \cite{kucina2022solution} investigated the reliability of cognitive conflict effects \hl{(as measured by response time)} in new versions of several standard cognitive-control tasks (e.g., Stroop, Flanker, Simon) that amplified cognitive interference via two task design features. First, they combined multiple sources of cognitive interference in the same task, for example by combining the Stroop and Simon effects to create a ``Stroopon'' task. Second, for a subset of trials, they required participants to make multiple responses based on both relevant and irrelevant stimuli attributes. Compared to previous versions of these tasks, these manipulations had the effect of increasing task demands, which resulted in greater between-participants variance and, consequently, \hl{required hundreds fewer trials to achieve a reliability of $\rho = 0.8$.} (See also Snijder and colleagues \cite{snijder2022psychometric} for a similar redesign of classic cognitive-control tasks that improved reliability in part by increasing proactive control demands.)

A related strategy is to calibrate the difficulty of the task to the average ability of the population of interest. For example, consider a task trial with two response options. Given the Bernoulli distribution, the variance in responses on this trial will be maximal when the probability of choosing either response is equal. Thus, aggregating across many trials, between-participants variance is maximized --- and task reliability is improved --- when the difficulty of all items is matched to the average ability of the sample \cite{gulliksen1945relation} (or slightly higher if participants can guess the correct response \cite{lord1952relation, feldt1993relationship}). Of course, this design principle is only helpful to the degree that a researcher knows the average ability level of their participants. If this is unknown or poorly characterized, then it is instead desirable to design a task with trials spanning a range of difficulty levels.

@@

A final strategy for increasing between-participant variance is to simply recruit more diverse samples. While convenient, undergraduate students from a single university are likely to be relatively homogeneous in their cognitive profiles. It may be worthwhile instead to recruit participants from the community or from an online labor market (e.g., Amazon Mechanical Turk, Prolific Academic, CloudResearch Panels). With regard to the latter, because online participants typically complete experiments from their homes or other poorly-controlled environments, they are more likely to be distracted or to multi-task during an experiment \cite{newman2021data}. Thus, when recruiting online samples, experimenters should take special care to ensure that an increase in between-participants variance is not offset by a concomitant increase in measurement noise. \hl{Separately, online participants may be more familiar with particular experimental paradigms due to previous exposure} \cite{chandler2014nonnaivete}, \hl{which may attenuate between-participants performance variability for the reasons previously mentioned (e.g., practice effects). Thus, researchers running experiments online may want to alter task paradigms so that they appear less similar to preexisting versions and/or limit the recruitment of highly-experienced participants} \cite{robinson2019tapped}.

@@

We thank Rachel Bedder and Felix Jan Nitsch for helpful feedback on this manuscript. This project was made possible with support from the National Center for Advancing Translational Sciences (NCATS), a component of the National Institute of Health (NIH), under award number UL1TR003017a and a National Science Foundation Graduate Research Fellowship.
